Pyg4ometry: a python library for the creation of Monte Carlo radiation transport physical geometries

### 2019-12-01 public/paper/pyg4ometry.tex
Updated paper

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%\documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage[T1]{fontenc}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, so called radiation transport (RT) programmes, these 
include MCNP \cite{Mcnp_Werner}, FLUKA \cite{Fluka_Ferrari,Fluka_Bohlen}, Geant3 \cite{Geant3_Brun} and Geant4 \cite{Geant4_Agostinelli}. 
All these codes are based on the Monte Carlo technique but each code either has a particular specialism or target user community.  RT Monte Carlo (RTMC) simulations 
have diverse uses, shielding calculations for radiological protection, detector performance, medical imaging and therapy and space radiation environment 
simulation are some examples. A fundamental requirement of all of the codes is to supply a computer readable description of the physical geometry that 
the particles are passing through.  The creation of geometry files is typically a very time consuming activity and the simulation validity and performance is directly 
dependent on the quality of the geometry. The requirement of a geometry system that allows the conversion between files prepared for different codes will 
enable cross checks of the physics processes in different particle transport codes. Parallel to the creation of geometry files for PT programs, usually computer assisted 
design (CAD) files exist for systems which need to be simulated in PT programmes. The fundamental geometric representations in CAD files is typically not 
amenable for PT programmes.  For these reasons it is advantageous to create a software tool that allows particle transport code users to rapidly develop 
geometry files, convert between common PT programme geometry formats and finally import CAD models.

@@

\subsection{Constructive solid geometry (CSG)}
Constructive solid geometry uses boolean operations to model complex surfaces which represent a solid. Typically simple finite or infinite solids are combined 
with operations, such as union, intersection and difference to model more complex solids. FLUKA uses CSG to model solids.

\subsection{Boundary representation (BREP)}
Boundary representation consists of two parts, topology and geometry. Topological elements are faces, edges and vertices. No current PT applications use native 
file formats employed by CAD systems. The conversion of CAD BREP formats for loading in PT applications is typically performed via a tessellated format.

\subsection{Tessellated polygons}
Solid volumes can be defined using triangular, quadrilateral or tetrahedral meshes. Numerous formats exist to describe meshes, the ubiquitous being STL and more 
modern variants being PLY, OBJ, etc.

@@

\begin{figure*}[hbt]
  \normalsize
  \centering
  \includegraphics[width=0.7\textwidth]{./diagrams/workflow.pdf}
  \caption{\label{fig:workflow}Workflow}
\end{figure*}

@@

\subsection{Tessellation of solids}
For each Geant4 primitive solid a triangular tessellated mesh is generated and cached. This mesh is then used to determine the extent of placed instances of 
geometry (physical volumes) and meshes for CSG derived solids. CSG mesh calculations are performed using a BSP tree technique. Triangular meshes based
on CSG operations involving curved surfaces often contain large numbers of thin triangles. Before meshes are visualised or written to disk various algorithms from 
TetGEN can be employed to give the meshes more desirable features.

\subsection{Replica, parametrised and looped volumes}
A powerful feature of Geant4 and hence GDML is the ability to either repeat or parametrise geometry. The create mulitple replicas of a volume in a cartesian, cylindrical or 
spherical grid is known as a ReplicaVolume.

@@

@@

\subsection{Adding properties to volumes}
A pure geometry description is an insufficient description for a PT simulation. Often other attributes need to be assigned for example material definition, magnetic field or

@@

\section{Rapid geometry modelling}
Given the python scripting interface, expression  and tessellation engines. It is possible for a user to rapidly specify the geometrical layout of the PT problem, vary 
the parameters of the geometry and visualise the geometry. Finally when the desired result without geometry overlaps is finalised a GDML file can be output
from the internal memory representation.

@@

\section{FLUKA to Geant4 conversion}
FLUKA geometry is based upon a limited set of primitive volumes which can be combined using logical operations. Many FLUKA primitives are infinite in extent, so 
to emulate FLUKA geometry construction large but finite extent Geant4 primitive solids are used. Table XXX shows the mapping between FLUKA and Geant4 objects.
An ANTLR based expression parser is used to load and comprehend the FLUKA input syntax, an example is shown in XXX. Each FLUKA body is matched to a Geant4 
solid and a FLUKA region is represented by a Geant4 CSG tree is formed using the pyg4ometry classes.

\begin{table}[hbt!]
\centering
\begin{tabular}{|l|l|} \hline
FLUKA body							& pyg4ometry class 		\\ \hline
RPP 	(Rectangular parallelepiped)			& Box				\\
BOX	(General rectangular parallelepiped)		& Box				\\
SPH 	(Sphere)							& Orb				\\
RCC (Right circular cylinder)				& Cone				\\
REC (Right elliptical cylinder)				& EllipticalTube			\\
TRC (Truncated Right Angle Cone)			& Cone				\\
ELL (Ellipsoid of Revolution) 				& Ellipsoid				\\
WED/RAW (Right Angle Wedge)			& ExtrudedSolid		\\
ARB	(Arbitrary Convex Polyhedron)			& TessellatedSolid		\\
XYP 	($X$-$Y$ Infinite half-space)			& Box				\\
XZP 	($X$-$Z$ Infinite half-space)			& Box				\\
YZP 	($Y$-$Z$ Infinite half-space)			& Box				\\
PLA (Generic infinite half-space)			& Box				\\
XCC ($X$-axis Infinite Circular Cylinder)		& Tubs				\\
YCC ($Y$-axis Infinite Circular Cylinder)		& Tubs				\\
ZCC 	($Z$-axis Infinite Circular Cylinder)		& Tubs				\\
XEC 	($X$-axis Infinite Elliptical Cylinder)		& EllipticalTube			\\
YEC 	($Y$-axis Infinite Elliptical Cylinder)		& EllipticalTube			\\
ZEC ($Z$-axis Infinite Elliptical Cylinder)		& EllipticalTube			\\
QUA (Quadric surface) 					& TessellatedSolid		\\ \hline
\end{tabular}
\caption{Fluka geometry directives pyg4ometry classes.}
\end{table}

@@

\section{Geant4 to FLUKA conversion}

\begin{table}[hbt!]
\centering
\begin{tabular}{| l | l | } \hline
Geant4 solid			& FLUKA constuction			\\ \hline
Box					& XYP, XZP, YZP 				\\
Tube					& ZCC, XYP	 				\\
CutTube				& ZCC, PLA					\\
Cone				& TRC						\\
Para					& ARB						\\
Trd					& ARB						\\
Trap					& ARB						\\
Sphere				& SPH, WED					\\
Orb					& SPH						\\
Torus				& ZCC, PLA					\\
Polycone				& ARB						\\
GenericPolycone		& ARB						\\
Polyhedra				& ARB						\\
Eltube				& ZEC, XYP					\\
Ellipsoid				& ELL						\\
Elcone				& QUA						\\
Paraboloid			& QUA						\\
Hype					& QUA						\\
Tet					& PLA						\\
Xtru					& PLA						\\
TwistedBox			& ARB						\\
TwistedTtap			& ARB						\\
TwistedTrd			& ARB				 		\\
TwistedTube			& ARB						\\
Arb8					& ARB						\\
Tessellated			& N/A					 	\\
Union				& | zone1 | zone2				\\
Subtraction			& +zone1 -zone2				\\
Intersection			& +zone1 +zone2				\\
MultiUnion			& | zone1 | zone2 | zone3 | ...	\\
ScaledSolid			& geant4.solid.ScaledSolid		\\ \hline				
\end{tabular}
\caption{GDML/Geant4 solids and the mapping to FLUKA bodies}
\end{table}

@@

\subsection{Overlap checking in geometry}
Often small errors in geometry cause unexpected PT code behaviour.

@@

\section{Conclusions}
The authors believe that tools to quickly create, either ab-initio or by conversion geometry usable in particle transport Monte Carlo programmes will save significant 
amounts of user effort and ultimately yield more accurate simulations. Pyg4ometry is a relatively complete implementation of a geometry creation tool, whilst heavily 
internally based on Geant4 and GDML can have utility for users of all PT codes. Pyg4ometry can clearly be extended to other formats or applications. As for applications,
examples include  the  generation triangular mesh structures for GPU accelerated photon tracking in liquid noble  dark matter detectors and geometries for event visualisation. 
Although not implemented in the current version, a C++ output writer can be quickly implemented to generate geometry which can be compiled into a Geant4 application.

>>>>

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage[T1]{fontenc}
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, so called radiation transport (RT) programmes, these 
include MCNP \cite{Mcnp_Werner}, FLUKA \cite{Fluka_Ferrari,Fluka_Bohlen}, Geant3 \cite{Geant3_Brun} and Geant4 \cite{Geant4_Agostinelli}. 
All these codes are based on the Monte Carlo technique but each code either has a particular specialism or target user community.  RT Monte Carlo (RTMC) simulations 
have diverse uses, shielding calculations for radiological protection, detector performance, medical imaging and therapy and space radiation environment 
simulation are some examples. A fundamental requirement of all of the codes is to supply a computer readable description of the physical geometry that 
the particles are passing through.  The creation of geometry files is typically a very time consuming activity and the simulation validity and performance is directly 
dependent on the quality of the geometry. The requirement of a geometry system that allows the conversion between files prepared for different codes will 
enable cross checks of the physics processes in different particle transport codes. Parallel to the creation of geometry files for RT programs, usually computer assisted 
design (CAD) files exist for systems which need to be simulated in RT programmes. The fundamental geometric representations in CAD files is typically not 
amenable for RT programmes.  For these reasons it is advantageous to create a software tool that allows particle transport code users to rapidly develop 
geometry files, convert between common RT programme geometry formats and finally import CAD models.

@@

\subsection{Constructive solid geometry (CSG)}
Constructive solid geometry uses boolean operations (subtraction, intersection and union) between simple solids shapes (cube, cylinder, sphere etc) to model complex 
surfaces which represent a solid. Typically simple finite or infinite solids are combined with operations, such as union, intersection and difference to model more 
complex solids. FLUKA uses CSG to model solids.

\subsection{Boundary representation (BREP)}
Boundary representation consists of two parts, topology and geometry. Topological elements are faces, edges and vertices. No current RT applications use native 
file formats employed by CAD systems. The conversion of CAD BREP formats for loading in PT applications is typically performed via a tessellated format, although it 
is possible to decompose BREP descriptions to simple half-spaces and solids as used in CSG descriptions.

\subsection{Tessellated polygons}
Solid volumes can be defined using triangular, quadrilateral or tetrahedral meshes. Numerous formats exist to describe meshes, the ubiquitous being STL and more 
modern variants being PLY, OBJ, etc. For solids with curved faces a tessellated mesh will always give an approximate description. As the mesh deviation distance from the 
solid decreases the number of polygons increases and hence the memory consumption and execution time of the RTMC.

@@

\begin{figure*}[hbt]
  \normalsize
  \centering
  \includegraphics[width=0.7\textwidth]{./diagrams/workflow.pdf}
  \caption{\label{fig:workflow}Pyg4ometry geometry conversion workflow.}
\end{figure*}

@@

\subsection{Replica, division, parametrised and looped volumes}
A powerful feature of Geant4 and hence GDML is the ability to either repeat, divide or parametrise geometry. The create multiple replicas of a volume in a cartesian, cylindrical or 
spherical grid is known as a ReplicaVolume. Currently GDML loops are not supported by pyg4ometry and will be implemented in a future release.

@@

\subsection{Tessellation of solids (meshing)}
For each Geant4 primitive solid a triangular tessellated mesh is generated and cached. This mesh is then used to determine the extent of placed instances of 
geometry (physical volumes) and meshes for CSG derived solids. CSG mesh calculations are performed using a BSP tree technique. Triangular meshes based
on CSG operations involving curved surfaces often contain large numbers of thin triangles. Before meshes are visualised or written to disk various algorithms from 
TetGEN can be employed to give the meshes more desirable features.

@@

\subsection{Adding properties to volumes}
A pure geometry description is an insufficient description for a RT simulation. Often other attributes need to be assigned for example material definition, magnetic field or  
optical surface properties.

@@

\section{Rapid geometry modelling}
Given the python scripting interface, expression  and tessellation engines. It is possible for a user to rapidly specify the geometrical layout of the RT problem, vary 
the parameters of the geometry and visualise the geometry. Finally when the desired result without geometry overlaps is finalised a GDML file can be output
from the internal memory representation.

@@

\section{FLUKA to Geant4 conversion}
FLUKA geometry is based upon a limited set of primitive volumes which can be combined using logical operations. Many FLUKA primitives are infinite in extent, so 
to emulate FLUKA geometry construction large but finite extent Geant4 primitive solids are used. Table \ref{tab:Fluka2Geant4} shows the mapping between FLUKA and Geant4 objects.
An ANTLR based expression parser is used to load and comprehend the FLUKA input syntax, an example is shown in XXX. Each FLUKA body is matched to a Geant4 
solid and a FLUKA region is represented by a Geant4 CSG tree is formed using the pyg4ometry classes.

\begin{table}[hbt!]
\centering
\begin{tabular}{|l|l|} \hline
FLUKA body							& pyg4ometry class 		\\ \hline
RPP 	(Rectangular parallelepiped)			& Box				\\
BOX	(General rectangular parallelepiped)		& Box				\\
SPH 	(Sphere)							& Orb				\\
RCC (Right circular cylinder)				& Cone				\\
REC (Right elliptical cylinder)				& EllipticalTube			\\
TRC (Truncated Right Angle Cone)			& Cone				\\
ELL (Ellipsoid of Revolution) 				& Ellipsoid				\\
WED/RAW (Right Angle Wedge)			& ExtrudedSolid		\\
ARB	(Arbitrary Convex Polyhedron)			& TessellatedSolid		\\
XYP 	($X$-$Y$ Infinite half-space)			& Box				\\
XZP 	($X$-$Z$ Infinite half-space)			& Box				\\
YZP 	($Y$-$Z$ Infinite half-space)			& Box				\\
PLA (Generic infinite half-space)			& Box				\\
XCC ($X$-axis Infinite Circular Cylinder)		& Tubs				\\
YCC ($Y$-axis Infinite Circular Cylinder)		& Tubs				\\
ZCC 	($Z$-axis Infinite Circular Cylinder)		& Tubs				\\
XEC 	($X$-axis Infinite Elliptical Cylinder)		& EllipticalTube			\\
YEC 	($Y$-axis Infinite Elliptical Cylinder)		& EllipticalTube			\\
ZEC ($Z$-axis Infinite Elliptical Cylinder)		& EllipticalTube			\\
QUA (Quadric surface) 					& TessellatedSolid		\\ \hline
\end{tabular}
\caption{FLUKA geometry directives pyg4ometry classes.} \label{tab:Fluka2Geant4}
\end{table}

@@

\section{Geant4 to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each of the Geant4 solids can be mapped to a FLUKA CSG region.
Some solids in Geant4 directly map to a FLUKA body, others need a construction of a simple FLUKA CSG tree.

\begin{table}[hbt!]
\centering
\begin{tabular}{| l | l | } \hline
Geant4 solid			& FLUKA region constuction		\\ \hline
Box					& +2 XYP + 2 XZP + 2 YZP 		\\
Tube					& +ZCC - 2 XYP	 			\\
CutTube				& +ZCC - 2 PLA				\\
Cone				& +TRC						\\
Para					& +ARB						\\
Trd					& +ARB						\\
Trap					& +ARB						\\
Sphere				& +SPH - SPH  - WED - TRC		\\
Orb					& +SPH						\\
Torus				& +$N$ ZCC  - $N$ PLA			\\
Polycone				& +$N$ ARB					\\
GenericPolycone		& +$N$ ARB					\\
Polyhedra				& +$N$ ARB					\\
Eltube				& +ZEC  - 2 XYP				\\
Ellipsoid				& +ELL - 2 XYP		 			\\
Elcone				& +QUA - 2 XYP				\\
Paraboloid			& +QUA - 2 XYP				\\
Hype					& +QUA - 2 XYP				\\
Tet					& +ARB						\\
Xtru					& +$N$ ARB					\\
TwistedBox			& +$N$ ARB					\\
TwistedTtap			& +$N$ ARB					\\
TwistedTrd			& +$N$ ARB				 	\\
TwistedTube			& +$N$ ARB					\\
Arb8					& +ARB						\\
Tessellated			& N/A					 	\\
Union				& | zone1 | zone2				\\
Subtraction			& +zone1 -zone2				\\
Intersection			& +zone1 +zone2				\\
MultiUnion			& | zone1 | zone2 | zone3 | ...	\\
ScaledSolid			& geant4.solid.ScaledSolid		\\ \hline				
\end{tabular}
\caption{GDML/Geant4 solids and the mapping to FLUKA bodies}
\end{table}

@@

\subsection{Overlap checking in geometry}
Often small errors in geometry cause unexpected RT code behaviour.

@@

\section{Conclusions}
The authors believe that tools to quickly create, either ab-initio or by conversion geometry usable in particle transport Monte Carlo programmes will save significant 
amounts of user effort and ultimately yield more accurate simulations. Pyg4ometry is a relatively complete implementation of a geometry creation tool, whilst heavily 
internally based on Geant4 and GDML can have utility for users of all RT codes. Pyg4ometry can clearly be extended to other formats or applications. As for applications,
examples include  the  generation triangular mesh structures for GPU accelerated photon tracking in liquid noble  dark matter detectors and geometries for event visualisation. 
Although not implemented in the current version, a C++ output writer can be quickly implemented to generate geometry which can be compiled into a Geant4 application.
